Investigating the incidence and potential risk factors for hypertension-related mortalities in adults in the United States.

### 2025-04-11 code/analysis-code/statistical-analysis-v2-slim.qmd
Changes to Part 4

@@

```{r}
# Define file paths for saving/loading tuning results
tuned_path     <- here::here("results", "large-files", "rf_tuned.rds")
metrics_path   <- here::here("results", "output", "rf_tuned_metrics.rds")
best_param_path <- here::here("results", "output", "rf_best_params.rds")

# Check if tuning results already exist
if (file.exists(tuned_path) && file.exists(metrics_path) && file.exists(best_param_path)) {
  message( "Tuning results already saved. Loading RF tuning results...")

  # Load saved objects
  rf_tuned <- readRDS(tuned_path)
  rf_tuned_metrics <- readRDS(metrics_path)
  best_rf_params <- readRDS(best_param_path)

} else {
  message(" Running Random Forest tuning...")
# Set up a small tuning grid for efficiency
rf_grid <- grid_regular(
  mtry(range = c(1, 4)),     # Explore 1 to 4 predictors
  min_n(range = c(5, 25)),   # Minimum observations per leaf
  levels = 3)                # Total levels in grid for each param

# Tune the random forest model
set.seed(123)
rf_tuned <- tune_grid(
  rf_workflow,
  resamples = cv_folds,
  grid = rf_grid,
  metrics = metric_set(rmse, rsq, mae),
  control = control_grid(save_pred = TRUE))

# Collect metrics from the tuning process
rf_tuned_metrics <- collect_metrics(rf_tuned)
best_rf_params <- select_best(rf_tuned, metric = "rmse")

# Save results to avoid re-running this chunk
  saveRDS(rf_tuned, tuned_path)
  saveRDS(rf_tuned_metrics, metrics_path)
  saveRDS(best_rf_params, best_param_path)

  message("Tuning complete and results saved.")}

```
```{r}
## ---- fit_or_load_rf_model, message=TRUE, warning=FALSE -------------------

use_slim_model <- FALSE # Toggle this to FALSE to regenerate slim model if needed

slim_rf_path <- here::here("results", "output", "rf_model_minimal.rds")
full_rf_path <- here::here("results", "large-files", "rf_tuned.rds")
full_rf_metrics_path <- here::here("results", "output", "rf_tuned_metrics.rds")
best_rf_params_path <- here::here("results", "output", "rf_best_params.rds")

if (use_slim_model) {
  if (!file.exists(slim_rf_path)) {
    stop("Slimmed model not found. Please run with use_slim_model <- FALSE first to generate it.")
  }

  message("Loading slimmed RF model...")
  rf_slim <- readRDS(slim_rf_path)
  rf_final_model <- rf_slim$model
  rf_metrics <- rf_slim$metrics
  rf_vip <- rf_slim$vip
  best_rf_params <- readRDS(best_rf_params_path)

} else {
  if (!file.exists(full_rf_path)) {
    stop("Full Random Forest model not found. Please rerun the tuning block.")
  }

  message("Loading full RF model and generating slim version...")
  rf_tuned <- readRDS(full_rf_path)
  rf_metrics <- readRDS(full_rf_metrics_path)
  rf_final_model <- extract_workflow(rf_tuned)
  rf_vip <- extract_fit_parsnip(rf_final_model) %>% vip::vi()
  best_rf_params <- readRDS(best_rf_params_path)

  if (!file.exists(slim_rf_path)) {
    saveRDS(
      list(model = rf_final_model, metrics = rf_metrics, vip = rf_vip),
      file = slim_rf_path
    )
    message("Slimmed model saved.")}}

```
Creeating a slimmed down version of the tuning RDS file to reduce file size for Github and working with if statements to load this from the correct folder and to not overwrite files.

@@

@@

# Skip if already saved
if (file.exists(fit_path) && file.exists(pred_path)) {
  message("Final model and predictions already saved. Skipping re-run...")
  final_rf_fit <- readRDS(fit_path)
  final_rf_workflow <- readRDS(workflow_path)
  rf_predictions <- readRDS(pred_path)
  rf_metrics <- readRDS(metrics_path)

@@

  # Evaluate performance
  rf_metrics <- rf_predictions %>%
    metrics(truth = mortalities, estimate = .pred)

  # Save all results
  saveRDS(final_rf_fit, file = fit_path)
  saveRDS(final_rf_workflow, file = workflow_path)
  saveRDS(rf_predictions, file = pred_path)
  saveRDS(rf_metrics, file = metrics_path)

@@

# Review metrics
print(rf_metrics)

@@

# Load saved RF metrics
rf_metrics <- readRDS(here::here("results", "output", "rf_test_metrics.rds"))

# Convert to wide format for easy table summary
rf_summary_table <- rf_metrics %>%
  select(.metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate) %>%
  mutate(
    Model = "Random Forest",
    Notes = "Tuned with 4 predictors (mtry = 4), min_n = 15"
  ) %>%
  select(Model, RMSE = rmse, R2 = rsq, MAE = mae, Notes)

@@

```{r}
## Tune LASSO penalty parameter

# File paths
lasso_tuned_path <- here::here("results", "large-files", "lasso_tuned.rds")
lasso_metrics_path <- here::here("results", "output", "lasso_tuned_metrics.rds")
lasso_best_param_path <- here::here("results", "output", "lasso_best_params.rds")

# Load or tune
if (file.exists(lasso_tuned_path) && file.exists(lasso_metrics_path) && file.exists(lasso_best_param_path)) {
  message("Loading saved LASSO tuning results...")

  lasso_tuned <- readRDS(lasso_tuned_path)
  lasso_tuned_metrics <- readRDS(lasso_metrics_path)
  best_lasso_params <- readRDS(lasso_best_param_path)
} else {
  message("Running LASSO tuning...")

  lasso_grid <- grid_regular(penalty(range = c(-4, -1)), levels = 10)  # log10 scale

  set.seed(123)
  lasso_tuned <- tune_grid(
    lasso_workflow,
    resamples = cv_folds,
    grid = lasso_grid,
    metrics = metric_set(rmse, rsq, mae),
    control = control_grid(save_pred = TRUE))

  lasso_tuned_metrics <- collect_metrics(lasso_tuned)
  best_lasso_params <- select_best(lasso_tuned, metric = "rmse")

  # Save results
  saveRDS(lasso_tuned, lasso_tuned_path)
  saveRDS(lasso_tuned_metrics, lasso_metrics_path)
  saveRDS(best_lasso_params, lasso_best_param_path)

  message("LASSO Tuning complete and results saved.")
}

# Preview metrics
print(lasso_metrics)
```

Creeating a slimmed down version of the tuning RDS file to reduce file size for Github and working with if statements to load this from the correct folder and to not overwrite files.
```{r}
## ---- slim_lasso_model_save, message=TRUE, warning=FALSE -------------------

slim_lasso_path <- here::here("results", "output", "lasso_model_minimal.rds")

if (!file.exists(slim_lasso_path)) {
  message("Creating slimmed LASSO model object...")

  lasso_final <- extract_workflow(lasso_tuned)
  lasso_metrics <- lasso_tuned_metrics

  # Extract coefficients
  lasso_fit <- extract_fit_parsnip(lasso_final)
  lasso_coeffs <- as.matrix(lasso_fit$fit$beta)

  saveRDS(
    list(model = lasso_final, metrics = lasso_metrics, coefficients = lasso_coeffs),
    file = slim_lasso_path
  )

  message("Saved slimmed LASSO model as lasso_model_minimal.rds")
} else {
  message("Slimmed LASSO model already exists. Skipping save.")
}

The slim model can be toggled on or off here by changing from TRUE to FALSE if wanting to use the full version. The full RDS file is saved in the large-files folder which will not automatically upload to github.
```{r}
## ---- load_lasso_model, message=TRUE, warning=FALSE -------------------

use_slim_lasso_model <- TRUE  # Toggle to FALSE to use full model

if (use_slim_lasso_model) {
  if (!file.exists(here::here("results", "output", "lasso_model_minimal.rds"))) {
    stop("Slim LASSO model not found. Please run the slim-save chunk or disable slim mode.")
  }

  message("Loading slimmed LASSO model...")
  lasso_slim <- readRDS(here::here("results", "output", "lasso_model_minimal.rds"))

  lasso_final_model <- lasso_slim$model
  lasso_metrics <- lasso_slim$metrics
  lasso_coeffs <- lasso_slim$coefficients
  best_lasso_params <- readRDS(here::here("results", "output", "lasso_best_params.rds"))

} else {
  if (!file.exists(here::here("results", "large-files", "lasso_tuned.rds"))) {
    stop("Full LASSO model not found.")
  }

  message("Loading full LASSO model...")
  lasso_tuned <- readRDS(here::here("results", "large-files", "lasso_tuned.rds"))
  lasso_metrics <- readRDS(here::here("results", "output", "lasso_tuned_metrics.rds"))
  lasso_final_model <- extract_workflow(lasso_tuned)

  lasso_fit <- extract_fit_parsnip(lasso_final_model)
  lasso_coeffs <- as.matrix(lasso_fit$fit$beta)

  best_lasso_params <- readRDS(here::here("results", "output", "lasso_best_params.rds"))}

@@

@@

} else {
  # Finalize model
  final_lasso_model <- finalize_model(lasso_spec, best_lasso_params)

  # Build workflow
  final_lasso_workflow <- workflow() %>%
    add_model(final_lasso_model) %>%
    add_recipe(mortality_recipe)

@@

# Load test metrics
rf_metrics <- readRDS(here::here("results", "output", "rf_test_metrics.rds"))
lasso_metrics <- readRDS(here::here("results", "output", "lasso_test_metrics.rds"))

# Add model names
rf_metrics$model <- "Random Forest"
lasso_metrics$model <- "LASSO Regression"

# Combine
all_metrics <- bind_rows(rf_metrics, lasso_metrics)

@@

```{r}
model_perf_numeric <- bind_rows(rf_metrics, lasso_metrics) %>%
  select(model, .metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate)

>>>>

```
These plots are best viewed as images outside of R due to size.

@@

Creeating a slimmed down version of the tuning RDS file to reduce file size for Github and working with if statements to load this from the correct folder and to not overwrite files.

```{r}
## ---- rf_tune_and_fit_if_guarded, message=TRUE, warning=FALSE -------------------

library(tidymodels)
library(vip)
library(here)

# Create directories
dir.create(here("results", "output"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("results", "large-files"), recursive = TRUE, showWarnings = FALSE)

# Define output paths
rf_tuned_path     <- here("results", "large-files", "rf_tuned.rds")
rf_metrics_path   <- here("results", "output", "rf_tuned_metrics.rds")
rf_best_param_path <- here("results", "output", "rf_best_params.rds")
rf_grid_path      <- here("results", "output", "rf_grid.rds")
rf_slim_path      <- here("results", "output", "rf_model_minimal.rds")

# Check if all exist
if (file.exists(rf_tuned_path) && file.exists(rf_metrics_path) &&
    file.exists(rf_best_param_path) && file.exists(rf_slim_path)) {

  message("✅ All RF files found. Loading saved versions...")

  rf_tuned <- readRDS(rf_tuned_path)
  rf_metrics <- readRDS(rf_metrics_path)
  best_rf_params <- readRDS(rf_best_param_path)
  rf_slim <- readRDS(rf_slim_path)
  rf_final_model <- rf_slim$model
  rf_vip <- rf_slim$vip

} else {
  message("🔁 Fitting Random Forest from scratch...")

  # Define grid
  rf_grid <- grid_regular(
    mtry(range = c(1, 4)),
    min_n(range = c(5, 25)),
    levels = 3
  )
  saveRDS(rf_grid, rf_grid_path)

  # Tune
  set.seed(123)
  rf_tuned <- tune_grid(
    rf_workflow,
    resamples = cv_folds,
    grid = rf_grid,
    metrics = metric_set(rmse, rsq, mae),
    control = control_grid(save_pred = TRUE)
  )

  # Extract results
  rf_metrics <- collect_metrics(rf_tuned)
  best_rf_params <- select_best(rf_tuned, metric = "rmse")

  # Save tuning results
  saveRDS(rf_tuned, rf_tuned_path)
  saveRDS(rf_metrics, rf_metrics_path)
  saveRDS(best_rf_params, rf_best_param_path)

  # Finalize and fit full model
  final_rf_workflow <- finalize_workflow(rf_workflow, best_rf_params)

  set.seed(123)
  rf_final_model <- fit(final_rf_workflow, data = train_data)

  # VIP
  rf_vip <- extract_fit_parsnip(rf_final_model) %>% vip::vi()

  # Save slim model
  saveRDS(
    list(model = rf_final_model, metrics = rf_metrics, vip = rf_vip),
    file = rf_slim_path
  )

  message("✅ RF slim model and all outputs saved.")}

@@

```{r}
## ---- temporary_rf_runtime_log, message=TRUE, warning=FALSE -------------------

# Optional: Log run time and session info for audit trail
start_time <- Sys.time()
Sys.sleep(1)  # simulate runtime block
end_time <- Sys.time()

writeLines(as.character(end_time - start_time), here::here("results", "output", "rf_runtime_duration.txt"))
writeLines(capture.output(sessionInfo()), here::here("results", "output", "rf_session_info.txt"))

message("⏱️ Runtime and session info saved.")

@@

# Skip if already saved
if (file.exists(fit_path) && file.exists(pred_path)) {
  message("Final model and predictions already saved. Skipping re-run...")
  final_rf_fit <- readRDS(fit_path)
  final_rf_workflow <- readRDS(workflow_path)
  rf_predictions <- readRDS(pred_path)
  rf_test_metrics <- readRDS(metrics_path)

@@

  # Evaluate performance
  rf_test_metrics <- rf_predictions %>%
    metrics(truth = mortalities, estimate = .pred)

  # Save all results
  saveRDS(final_rf_fit, file = fit_path)
  saveRDS(final_rf_workflow, file = workflow_path)
  saveRDS(rf_predictions, file = pred_path)
  saveRDS(rf_test_metrics, file = metrics_path)

@@

# Review metrics
print(rf_test_metrics)

@@

# Load saved RF metrics
rf_test_metrics <- readRDS(here::here("results", "output", "rf_test_metrics.rds"))

# Convert to wide format for easy table summary
rf_summary_table <- rf_test_metrics %>%
  select(.metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate) %>%
  mutate(
    Model = "Random Forest",
    Notes = "Tuned with 4 predictors (mtry = 4), min_n = 15"
  ) %>%
  select(Model, RMSE = rmse, R2 = rsq, MAE = mae, Notes)

@@

Creeating a slimmed down version of the tuning RDS file to reduce file size for Github and working with if statements to load this from the correct folder and to not overwrite files.

```{r}
## ---- lasso_tune_and_fit_if_guarded, message=TRUE, warning=FALSE -------------------

library(tidymodels)
library(vip)
library(here)

# Create folders
dir.create(here::here("results", "output"), showWarnings = FALSE, recursive = TRUE)
dir.create(here::here("results", "large-files"), showWarnings = FALSE, recursive = TRUE)

# Define output paths
lasso_tuned_path        <- here("results", "large-files", "lasso_tuned.rds")
lasso_metrics_path      <- here("results", "output", "lasso_tuned_metrics.rds")
lasso_best_param_path   <- here("results", "output", "lasso_best_params.rds")
lasso_grid_path         <- here("results", "output", "lasso_grid.rds")
slim_lasso_path         <- here("results", "output", "lasso_model_minimal.rds")

# Check if all saved files already exist
if (file.exists(lasso_tuned_path) &&
    file.exists(lasso_metrics_path) &&
    file.exists(lasso_best_param_path) &&
    file.exists(slim_lasso_path)) {

  message("✅ All LASSO files found. Loading saved versions...")

  lasso_tuned <- readRDS(lasso_tuned_path)
  lasso_metrics <- readRDS(lasso_metrics_path)
  best_lasso_params <- readRDS(lasso_best_param_path)
  lasso_slim <- readRDS(slim_lasso_path)
  final_lasso_model <- lasso_slim$model
  lasso_vip <- lasso_slim$vip
  lasso_coeffs <- lasso_slim$coefficients

} else {
  message("🔁 Fitting LASSO model from scratch...")

  # Define grid
  lasso_grid <- grid_regular(penalty(range = c(-4, 0)), levels = 20)
  saveRDS(lasso_grid, lasso_grid_path)

  # Tune
  set.seed(123)
  lasso_tuned <- tune_grid(
    lasso_workflow,
    resamples = cv_folds,
    grid = lasso_grid,
    metrics = metric_set(rmse, rsq, mae),
    control = control_grid(save_pred = TRUE)
  )

  # Extract and save tuning results
  lasso_metrics <- collect_metrics(lasso_tuned)
  best_lasso_params <- select_best(lasso_tuned, metric = "rmse")

  saveRDS(lasso_tuned, lasso_tuned_path)
  saveRDS(lasso_metrics, lasso_metrics_path)
  saveRDS(best_lasso_params, lasso_best_param_path)

  # Finalize and fit
  final_lasso_workflow <- finalize_workflow(lasso_workflow, best_lasso_params)

  set.seed(123)
  final_lasso_model <- fit(final_lasso_workflow, data = train_data)

  # Extract coefficients
  lasso_coeffs <- as.matrix(extract_fit_parsnip(final_lasso_model)$fit$beta)

  # VIP table
  lasso_vip <- tibble(
    variable = rownames(lasso_coeffs),
    importance = abs(lasso_coeffs[, 1])
  ) %>% arrange(desc(importance))

  # Save slimmed version
  saveRDS(
    list(model = final_lasso_model, metrics = lasso_metrics, coefficients = lasso_coeffs, vip = lasso_vip),
    file = slim_lasso_path
  )

  message("✅ Final LASSO model fitted and slim version saved.")}
```

The slim model can be toggled on or off here by changing from TRUE to FALSE if wanting to use the full version. The full RDS file is saved in the large-files folder which will not automatically upload to github.
```{r}
## ---- lasso_confidence_metrics, message=TRUE, warning=FALSE -------------------

lasso_rmse_summary <- lasso_metrics %>%
  filter(.metric == "rmse") %>%
  mutate(
    lower_95 = mean - 1.96 * std_err,
    upper_95 = mean + 1.96 * std_err
  ) %>%
  arrange(mean)

write.csv(lasso_rmse_summary, here::here("results", "output", "lasso_rmse_summary.csv"), row.names = FALSE)

message("✅ LASSO RMSE summary with CI estimates saved.")
```

@@

```{r}
## ---- temporary_lasso_runtime_log, message=TRUE, warning=FALSE -------------------

start_time <- Sys.time()
Sys.sleep(1)  # Simulated delay
end_time <- Sys.time()

writeLines(as.character(end_time - start_time), here::here("results", "output", "lasso_runtime_duration.txt"))
writeLines(capture.output(sessionInfo()), here::here("results", "output", "lasso_session_info.txt"))

message("⏱️ LASSO runtime and session info saved.")

@@

} else {
  # Finalize model
  final_lasso_workflow <- finalize_workflow(lasso_spec, best_lasso_params)

@@

# Load test metrics
rf_test_metrics <- readRDS(here::here("results", "output", "rf_test_metrics.rds"))
lasso_metrics <- readRDS(here::here("results", "output", "lasso_test_metrics.rds"))

# Add model names
rf_test_metrics$model <- "Random Forest"
lasso_metrics$model <- "LASSO Regression"

# Combine
all_metrics <- bind_rows(rf_test_metrics, lasso_metrics)

@@

```{r}
model_perf_numeric <- bind_rows(rf_test_metrics, lasso_metrics) %>%
  select(model, .metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate)

### 2025-04-11 products/manuscript/Manuscript.qmd
Changes to Part 4

This data was downloaded from the CDC online database at the following
url:
https://data.cdc.gov/Heart-Disease-Stroke-Prevention/Rates-and-Trends-in-Hypertension-related-Cardiovas/uc9k-vc2j/about_data.
A csv file was downloaded which was almost 300MB in size. The initial
document/dataset contained 1103872 observations across 24 variables.
There were some initial processing steps performed to evaluate and
delete some of the unnecessary columns and to save this data to a .rds
document to reduce the size. This was all outlined in the
"processingcode" document, which will not run automatically because the
Excel (csv) file will not be available due to size concerns.

>>>>

This data was downloaded from the CDC online database at the following
url:
https://data.cdc.gov/Heart-Disease-Stroke-Prevention/Rates-and-Trends-in-Hypertension-related-Cardiovas/uc9k-vc2j/about_data.
A csv file was downloaded which was almost 300MB in size. The initial
document/dataset contained 1103872 observations across 24 variables.
There were some initial processing steps performed to evaluate and
delete some of the unnecessary columns and to save this data to a .rds
document to reduce the size. This was all outlined in the
"processingcode" document (and README file), which will not run automatically because the
Excel (csv) file will not be available due to size concerns.
